RegressionPvNP

### 2017-05-23 RegressionPvNP.Rmd
Starting to firm up analysis

---
title: "RegressionPvNP"
author: "Sarah Killcoyne"
date: "31 March 2017"
output: 
  html_document: 
    fig_caption: yes
    fig_height: 10
    fig_width: 10
    number_sections: yes
    toc: yes
---

@@

data.files = list.files(paste(data, 'QDNAseq',sep='/'), full.names=T)
plot.dir = paste(data, 'Analysis/multipcf_plots_fitted_perPatient', sep='/')

if (length(list.files(plot.dir)) <= 0)
  stop(paste("No analysis files found in", plot.dir ))

@@

```{r apclust-func, echo=F, message=F, warning=F}
apclust.data<-function(segdata, samples) {
  x1 = segdata[,intersect(colnames(segdata), samples)]

  if (length(samples) <= 1) { 
    x1 = as.data.frame(x1)
    colnames(x1) = samples
  } else {
    x1 = x1[, samples]
  }
  rownames(x1) = (segdata[,c(1:4)] %>%
                    rowwise() %>%
                    mutate(location=paste(paste(chrom, arm, sep=''), '.', start.pos, '-', end.pos, sep='')))$location
  return(x1)  
}
```

@@

```{r tile, echo=F, message=F, warning=F, include=F}

chr.lengths = get.chr.lengths()
chr.lengths$chrom = sub('chr','',chr.lengths$chrom)
chr.lengths$start = 1

tile.w=1e7

genome = makeGRangesFromDataFrame(chr.lengths[1:22,], seqnames.field = 'chrom', end.field='chr.length')
tiles = tile(genome, width=tile.w)

tile.sample.segments<-function(tileFile, sample.data) {
  if (file.exists(tileFile)) {
    message(paste("Loading", tileFile))
    load(tileFile, verbose=T)
  } else {
    message("Tiling patient data...")
    grList = lapply(sample.data, function(df) {
      print(unique(df$info$Patient))
      df$info = arrange(df$info, Endoscopy.Year, Pathology)

      x1 = apclust.data(df$seg.vals, df$info$Samplename)

      segnames = as.data.frame(do.call(rbind, sapply(rownames(x1), strsplit, '[p|q].|-')))
      colnames(segnames) = c('chr','start','end')
      segnames[c('start','end')] = lapply(segnames[c('start','end')], function(x) as.numeric(as.character(x)))

      pt = unique(df$info$Patient)
      return( makeGRangesFromDataFrame(cbind(segnames,pt,x1), keep.extra.columns=T) )
    })

    sampleNames = unlist(sapply(sample.data, function(df) {
      df$info = arrange(df$info, Endoscopy.Year, Pathology)
      x1 = apclust.data(df$seg.vals, df$info$Samplename)
      colnames(x1)
    }))

    tile.segments<-function(tiles, gr, mergedSegments) {
      ov = findOverlaps(tiles, gr)
      for (qh in unique(queryHits(ov))) {
        #print(qh)
        currentTile = tiles[[qh]]
        curov = findOverlaps(currentTile, gr)

        for (i in 1:length(curov)) {
          segment = currentTile[ queryHits(curov)[i]  ]
          #print(segment)
          rows = with(mergedDf, which( chr==as.character(seqnames(segment)) & start == start(segment) & end == end(segment)))
          segmentVals = as.data.frame(elementMetadata(gr[ subjectHits(curov)[i] ]))
          mergedSegments[rows, names(segmentVals)[-1]] = segmentVals[,-1]
        }
      }
      return(mergedSegments)  
    }

    mergedDf = do.call(rbind, lapply(tiles, function(tile) { 
      cbind('chr'=as.character(seqnames(tile)), as.data.frame(ranges(tile))[1:2]) 
    }) )
    mergedDf[,sampleNames] = NA

    for (gr in grList) {
      #print(unique(gr$pt))
      mergedDf = tile.segments(tiles, gr, mergedDf)
    }

    #tmp = mergedDf
    mergedDf[is.na(mergedDf)] = 0
    rownames(mergedDf) = with(mergedDf, paste(chr, ':', start, '-', end, sep=''))

    save(mergedDf, file=tileFile)
  }
  return(mergedDf)
}

mergedDf = tile.sample.segments(paste(as.character(tile.w),'_tiledpts.Rdata', sep=''), patient.data)
dim(mergedDf)
#validatioMergedDf = tile.sample.segments(paste(as.character(tile.w),'validation_tiledpts.Rdata', sep=''), validation.patient.data)

@@

```{r xvalpt, message=F, warning=F, echo=T, fig.height=16, fig.width=16}
coefs = list(); plots = list(); performance.at.1se = list()
folds = 10; splits = 5
alpha.values = c(0,0.5,0.7,0.8,0.9,1)
if (file.exists('/tmp/all.pt.alpha.Rdata')) {
  load('/tmp/all.pt.alpha.Rdata', verbose=F)
} else {
  for (a in alpha.values) {
    fit0 <- glmnet(dysplasia.df, labels, alpha=a, nlambda=1000, family='binomial') # all patients
    #autoplot(fit0, xvar='lambda', main=paste('fit0, all samples, alpha=',a,sep='')) + theme(legend.position='none')

@@

    performance.at.1se[[as.character(a)]] = subset(cv.patient$lambdas, lambda.1se == T)
  }
  save(plots, coefs, performance.at.1se, file='/tmp/all.pt.alpha.Rdata')
}

@@

ggplot(performance.at.1se, aes(alpha, mean)) + geom_point() +
  geom_errorbar(aes(ymin=mean-sme, ymax=mean+sme)) + ylim(0.5,0.8) +
  geom_text(aes(label=n.Coef), nudge_y=0.03) +
  geom_text(aes(label=paste('(',`75%`,')',sep='')), nudge_y=0.02) +
  labs(x='Elasticnet penalty value', y='Model performance at lambda.1se', 
       title=paste('All samples,', folds, 'folds,', splits, 'patient splits'))

@@

@@

@@

no.hgd.plots = list(); coefs = list(); performance.at.1se = list()
if (file.exists('/tmp/nohgd.Rdata')) {
  load('/tmp/nohgd.Rdata', verbose=T)
} else {
  # No HGD/IMC
  samples = intersect(rownames(dysplasia.df), subset(info, Pathology %nin% c('HGD', 'IMC'))$Samplename)
  for (a in alpha.values) {
    # all patients
    fitNoHGD <- glmnet(dysplasia.df[samples,], labels[samples], alpha=a, family='binomial', nlambda=1000)

@@

    performance.at.1se[[as.character(a)]] = subset(cv.nohgd$lambdas, lambda.1se == T)
  }
  save(no.hgd.plots, coefs, performance.at.1se, file='/tmp/nohgd.Rdata')
}

@@

ggplot(performance.at.1se, aes(alpha, mean)) + geom_point() +
  geom_errorbar(aes(ymin=mean-sme, ymax=mean+sme)) + ylim(0.5,0.8) +
  geom_text(aes(label=n.Coef), nudge_y=0.03) +
  geom_text(aes(label=paste('(',`75%`,')',sep='')), nudge_y=0.02) +
  labs(x='Elasticnet penalty value', y='Model performance at lambda.1se', 
       title=paste('No HGD/IMC samples,', folds, 'folds,', splits, 'patient splits'))

@@

```{r noLGD, echo=T, message=F, warning=F, fig.height=16, fig.width=16}
nolgd.plots = list(); coefs = list(); performance.at.1se = list()
if (file.exists('/tmp/nolgd.Rdata')) {
  load('/tmp/nolgd.Rdata', verbose=T)
} else {
  # No LGD
  samples = intersect(rownames(dysplasia.df), c(subset(info, Status == 'NP')$Samplename, subset(info, Pathology %nin% c('HGD', 'IMC', 'LGD') & Status == 'P')$Samplename))

@@

    performance.at.1se[[as.character(a)]] = subset(cv.nolgd$lambdas, lambda.1se == T)
  }
  save(nolgd.plots, coefs, performance.at.1se, file='/tmp/nolgd.Rdata')
}

do.call(grid.arrange, c(plots, top="No HGD/IMC/LGD samples", ncol=2))
```

@@

ggplot(performance.at.1se, aes(alpha, mean)) + geom_point() +
  geom_errorbar(aes(ymin=mean-sme, ymax=mean+sme)) + ylim(0.5,0.8) +
  geom_text(aes(label=n.Coef), nudge_y=0.03) +
  geom_text(aes(label=paste('(',`75%`,')',sep='')), nudge_y=0.02) +
  labs(x='Elasticnet penalty value', y='Model performance at lambda.1se', 
       title=paste('No HGD/IMC/LGD samples,', folds, 'folds,', splits, 'patient splits'))

@@

if (file.exists('/tmp/few.Rdata')) {
  load('/tmp/few.Rdata', verbose=T)
} else {
  # Few
  samples = c(nonps, as.vector(unlist(lapply(patient.data[subset(sum.patient.data, Status == 'P' & total.samples < m)$Patient], function(pt) pt$info$Samplename))))

@@

    few.plots[[as.character(a)]] = arrangeGrob(cv.back.few$plot+ggtitle('Classification'), cv.back.few$deviance.plot+ggtitle('Binomial Deviance'), top=paste('alpha=',a,sep=''), ncol=2)
  }
  save(few.plots, file='/tmp/few.Rdata')
} 
do.call(grid.arrange, c(few.plots, ncol=2, top=paste(nrow(subset(sum.patient.data, Status == 'P' & total.samples < m))," Progressors with < median # samples (",m,")")))

# Many
if (file.exists('/tmp/many.Rdata')) {
  load('/tmp/many.Rdata', verbose=T)
} else {
  samples = c(nonps, as.vector(unlist(lapply(patient.data[subset(sum.patient.data, Status == 'P' & total.samples >= m)$Patient], function(pt) pt$info$Samplename))))

@@

    many.plots[[as.character(a)]] = arrangeGrob(cv.back.many$plot+ggtitle('Classification'), cv.back.many$deviance.plot+ggtitle('Binomial Deviance'), top=paste('alpha=',a,sep=''), ncol=2)
  }
  save(many.plots, file='/tmp/many.Rdata')
}

@@

if (file.exists('/tmp/loo.Rdata')) {
  load('/tmp/loo.Rdata', verbose=T)
} else {
  performance.at.1se = c(); coefs = list(); plots = list(); fits = list()
  # Remove each patient (LOO)
  for (pt in names(pg.samp)) {
    tmp.patient.data = patient.data[subset(sum.patient.data, Patient != pt)$Patient]
    samples = as.vector(unlist(sapply(tmp.patient.data, function(df) df$info$Samplename )))

    train.rows = which(rownames(dysplasia.df) %in% samples)
    training = dysplasia.df[train.rows,]
    test = dysplasia.df[-train.rows,]

    # Predict function giving me difficulty when I have only a single sample, this ensures the dimensions are the same
    sparsed_test_data <- Matrix(data=0, nrow=ifelse(nrow(test) > 1, nrow(test), 1),  ncol=ncol(training),
                            dimnames=list(pg.samp[[pt]]$Samplename,colnames(training)), sparse=T)
    for(i in colnames(test)) sparsed_test_data[,i] = test[,i]

@@

      pm = predict(fitLOO, newx=sparsed_test_data, s=cv$lambda.1se, type='response')
      sy = sqrt(binomial.deviance(pm, labels[pg.samp[[pt]]$Samplename]))

@@

    } else {
      warning(paste("Patient", pt, "did not have a 1se"))
    }
  }
  save(plots, performance.at.1se, coefs, fits, pg.samp, file='/tmp/loo.Rdata')
}

@@

@@

np = sapply(all.predictions[subset(sum.patient.data, Status == 'NP')$Patient], function(x) round(sum(x)/length(x),2))
pg = sapply(all.predictions[subset(sum.patient.data, Status == 'P')$Patient], function(x) round(sum(x)/length(x),2))

@@

In progressors, `r round(sum(hgd,na.rm=T)/length(hgd[!is.na(hgd)]), 2)*100`% of the HGD samples were correctly predicted.

When predicting each sample individually `r length(pg[pg == 1])` patients were correctly predicted in all samples. The remaining patients were predicted in `r max(pg[pg != 1])*100`% or fewer samples.

`r pander( as.data.frame(sort(pg)), caption='Proportion of progressor patients samples correctly predicted')`

@@

In non-progressors `r length(which(np == 1))` patients were predicted to be non-progressors in _all_ of their samples. `r length(which(np >= 0.75))` were predicted in 75% or more of their samples. `r length(np[which(np < 0.75 )])` patients had less than 75% of their samples predicted as NP.

`r pander( as.data.frame(sort(np)), caption='Proportion of NP patients samples correctly predicted') `

```{r timepoints2, echo=T, message=F, warning=F}
# For all pre-HGD/IMC/LGD in progressors
pre.hgd = sapply(pg.samp[subset(sum.patient.data, Status == 'P')$Patient], function(pt) {
  e = subset(pt, Pathology %nin% c('HGD', 'IMC', 'LGD'))
  e$Prediction
  sum(as.integer(e$Prediction > 0.5))/nrow(e)
})

pred.by.endo<-function(pt, nyears) {
  e = subset(pt, PID %in% unique(pt$PID)[1:nyears])
  if (  length(which(e$Pathology %in% c('HGD','IMC'))) > 0 ) return(NA)
  p = as.vector(as.integer(e$Prediction > 0.5))
  names(p) = e$Endoscopy.Year
  return(p)
}

@@

p.predictions = as.data.frame(matrix(data=0,nrow=6, ncol=3, dimnames=list(c(), c('%.Predicted','n.Endoscopies','n.Patients'))))
for (i in 1:nrow(p.predictions)) {
  p = sapply(pg.samp[subset(sum.patient.data, Status == 'P')$Patient], pred.by.endo, i)
  pp = sapply(p, predicted.progression)
  pp = pp[!is.na(pp)]
  p.predictions[i,] =  rbind(round(sum(pp[!is.na(pp)]/length(pp[!is.na(pp)])), 2), i, length(pp))
}

ggplot(p.predictions, aes(y=`%.Predicted`,x=n.Endoscopies)) + 
  geom_bar(stat="identity", fill='dodgerblue3') + 
  geom_text(aes(label=paste(`%.Predicted`*100, '%', sep='')), nudge_y=-0.02) + 
  geom_text(aes(label=paste(n.Patients, 'patients')), nudge_y=0.005) + ggtitle("Progression predicted per endoscopy")
```

`r pander(p.predictions, justify='left', caption=paste("For any given endoscopy if a single sample (same pathology ID, different blocks) is predicted as 'progressor' that endoscopy is considered to be indicative of progression. At each endoscopy the percentage is calculated excluding any patients that are diagnosed HGD at that point. There are ",length(subset(sum.patient.data, Status == 'P')$Patient), " progressor patients. Based on that, ", p.predictions[1,1]*100, "% are correctly predicted at the first endoscopy.", sep=''))`

What about the non-progressors that had progressor predictions?
```{r, warning=F, message=F}

np.predictions = as.data.frame(matrix(data=0,nrow=6, ncol=3, dimnames=list(c(), c('%.Predicted','n.Endoscopies','n.Patients'))))
for (i in 1:nrow(np.predictions)) {
  p = sapply(pg.samp[subset(sum.patient.data, Status == 'NP')$Patient], pred.by.endo, i)
  pp = sapply(p, predicted.progression)
  pp = pp[!is.na(pp)]
  np.predictions[i,] =  rbind(round(sum(pp[!is.na(pp)]/length(pp[!is.na(pp)])), 2), i, length(pp))
}

@@

# How do I evaluate the predictions?

I'm a bit at a loss here. I understand why penalized regression wouldn't have a CI, but then I'm not sure how best to evaluate the predictions.  Do I just take the predictions at face value and leave it at that?

```{r, warning=F, message=F, fig.height=4, fig.width=4}
all = do.call(rbind,lapply(pg.samp, function(df) {
  df[,c('Status','Prediction','Prediction.Dev.Resid')]
}))

>>>>

---
title: "RegressionPvNP"
author: "Sarah Killcoyne"
date: "23 May 2017"
output: 
  html_document: 
    toc: yes
    toc_depth: 4
---

@@

data.files = list.files(paste(data, 'QDNAseq',sep='/'), full.names=T)
analysis.files = list.files(paste(data, 'Analysis', sep='/'), full.names=T)

@@

@@

```{r tile, echo=F, message=F, warning=F, include=F}
#tile.w=1e07
tile.w=5e06
tile.files = grep('tile_.*5e\\+06', analysis.files, value=T)
load(tile.files[1], verbose=T)
dim(mergedDf)

cache.dir = paste(data, 'Analysis/5e06', sep='/')
if (!dir.exists(cache.dir)) dir.create(cache.dir)

#validatioMergedDf = tile.sample.segments(paste(as.character(tile.w),'validation_tiledpts.Rdata', sep=''), validation.patient.data)

@@

```{r xvalpt, message=F, warning=F, echo=T, fig.height=16, fig.width=16}
coefs = list(); plots = list(); performance.at.1se = list()
folds = 10; splits = 5
alpha.values = c(0,0.5,0.7,0.8,0.9,1)

file = paste(cache.dir, 'all.pt.alpha.Rdata', sep='/')
if (file.exists(file)) {
  load(file, verbose=T)
} else {
  for (a in alpha.values) {
    fit0 <- glmnet(dysplasia.df, labels, alpha=a, nlambda=1000, family='binomial') # all patients
    #autoplot(fit0, xvar='lambda', main=paste('fit0, all samples, alpha=',a,sep='')) + theme(legend.position='none')

@@

    performance.at.1se[[as.character(a)]] = subset(cv.patient$lambdas, lambda.1se == T)
  }
  save(plots, coefs, performance.at.1se, file=file)
}

@@

ggplot(performance.at.1se, aes(alpha, mean)) + geom_point() + geom_text( aes(label=round(mean, 3)), hjust=-0.5) +
  geom_errorbar(aes(ymin=mean-sme, ymax=mean+sme)) + ylim(0.5,0.8) +
  geom_text(aes(label=n.Coef), nudge_y=0.03) +
  geom_text(aes(label=paste('(',`75%`,')',sep='')), nudge_y=0.02) +
  labs(x='Elasticnet penalty value', y='Model performance at lambda.1se', 
       title=paste('All samples,', folds, 'folds,', splits, 'patient splits'))

@@

pander(as.data.frame(sort(names(which((coef.stable[['1']]) >= 0.75)))), justify='left')
```

@@

### Feature Information

```{r, warning=F, message=F}
#topF = do.call(rbind, strsplit( names(which((coef.stable[['1']]) >= 0.75)), ':|-'))
#colnames(topF) = c('chr','start','end')
#topF = apply(topF, 2, as.numeric)

topF = names(which((coef.stable[['1']]) >= 0.75))

library(biomaRt)
mart = biomaRt::useMart("ENSEMBL_MART_ENSEMBL", host="grch37.ensembl.org")
biomart = useDataset("hsapiens_gene_ensembl", mart)

attr = c('ensembl_gene_id','hgnc_symbol','start_position','end_position', 'chromosome_name', 'go_id', 'entrezgene')
bm = getBM(mart=biomart, attributes=attr, filters=list('chromosomal_region' = topF))
bm = bm[-which(bm$hgnc_symbol == ""),] 
#head(bm)

bm2 = ddply(bm, .(entrezgene, ensembl_gene_id, hgnc_symbol), summarise, 'chrom'=unique(chromosome_name), 'start'=min(start_position), 'end'=max(end_position), 'GO.IDs'=paste(go_id, collapse=','))

ccgenes = read.table('~/Data/CosmicCensusGenes.tsv', sep='\t', header=T, stringsAsFactors=F)

bm2 = subset(bm2, hgnc_symbol %in% ccgenes$Gene.Symbol)

features.in.prog = dysplasia.df[names(which(labels == 1)),topF]

#features.in.prog <= 0.9
#features.in.prog >= 1.1

@@

no.hgd.plots = list(); coefs = list(); performance.at.1se = list()

file = paste(cache.dir, 'nohgd.Rdata', sep='/')
if (file.exists(file)) {
  load(file, verbose=T)
} else {
  # No HGD/IMC
  samples = intersect(rownames(dysplasia.df), subset(info, Pathology %nin% c('HGD', 'IMC'))$Samplename)
  for (a in alpha.values) {
    # all patients
    fitNoHGD <- glmnet(dysplasia.df[samples,], labels[samples], alpha=a, family='binomial', nlambda=1000)

@@

    performance.at.1se[[as.character(a)]] = subset(cv.nohgd$lambdas, lambda.1se == T)
  }
  save(no.hgd.plots, coefs, performance.at.1se, file=file)
}

@@

ggplot(performance.at.1se, aes(alpha, mean)) + geom_point() + geom_text( aes(label=round(mean, 2)), hjust=-0.5) +
  geom_errorbar(aes(ymin=mean-sme, ymax=mean+sme)) + ylim(0.5,0.8) + 
  geom_text(aes(label=n.Coef), nudge_y=0.03) +
  geom_text(aes(label=paste('(',`75%`,')',sep='')), nudge_y=0.02) +
  labs(x='Elasticnet penalty value', y='Model performance at lambda.1se', 
       title=paste('No HGD/IMC samples,', folds, 'folds,', splits, 'patient splits'))

@@

```{r noLGD, echo=T, message=F, warning=F, fig.height=16, fig.width=16}
file = paste(cache.dir, 'nolgd.Rdata', sep='/')

nolgd.plots = list(); coefs = list(); performance.at.1se = list()
if (file.exists(file)) {
  load(file, verbose=T)
} else {
  # No LGD
  samples = intersect(rownames(dysplasia.df), c(subset(info, Status == 'NP')$Samplename, subset(info, Pathology %nin% c('HGD', 'IMC', 'LGD') & Status == 'P')$Samplename))

@@

    performance.at.1se[[as.character(a)]] = subset(cv.nolgd$lambdas, lambda.1se == T)
  }
  save(nolgd.plots, coefs, performance.at.1se, file=file)
}

do.call(grid.arrange, c(nolgd.plots, top="No HGD/IMC/LGD samples", ncol=2))
```

@@

ggplot(performance.at.1se, aes(alpha, mean)) + geom_point() + geom_text( aes(label=round(mean, 2)), hjust=-0.5) +
  geom_errorbar(aes(ymin=mean-sme, ymax=mean+sme)) + ylim(0.5,0.8) +
  geom_text(aes(label=n.Coef), nudge_y=0.03) +
  geom_text(aes(label=paste('(',`75%`,')',sep='')), nudge_y=0.02) +
  labs(x='Elasticnet penalty value', y='Model performance at lambda.1se', 
       title=paste('No HGD/IMC/LGD samples,', folds, 'folds,', splits, 'patient splits'))

@@

file = paste(cache.dir, 'few.Rdata', sep='/')

if (file.exists(file)) {
  load(file, verbose=T)
} else {
  # Few
  samples = c(nonps, as.vector(unlist(lapply(patient.data[subset(sum.patient.data, Status == 'P' & total.samples < m)$Patient], function(pt) pt$info$Samplename))))

@@

    few.plots[[as.character(a)]] = arrangeGrob(cv.back.few$plot+ggtitle('Classification'), cv.back.few$deviance.plot+ggtitle('Binomial Deviance'), top=paste('alpha=',a,sep=''), ncol=2)
  }
  save(few.plots, file=file)
} 
do.call(grid.arrange, c(few.plots, ncol=2, top=paste(nrow(subset(sum.patient.data, Status == 'P' & total.samples < m))," Progressors with < median # samples (",m,")")))

# Many
file = paste(cache.dir, 'many.Rdata', sep='/')

if (file.exists(file)) {
  load(file, verbose=T)
} else {
  samples = c(nonps, as.vector(unlist(lapply(patient.data[subset(sum.patient.data, Status == 'P' & total.samples >= m)$Patient], function(pt) pt$info$Samplename))))

@@

    many.plots[[as.character(a)]] = arrangeGrob(cv.back.many$plot+ggtitle('Classification'), cv.back.many$deviance.plot+ggtitle('Binomial Deviance'), top=paste('alpha=',a,sep=''), ncol=2)
  }
  save(many.plots, file=file)
}

@@

file = paste(cache.dir, 'loo.Rdata', sep='/')

if (file.exists(file)) {
  load(file, verbose=T)
} else {
  performance.at.1se = c(); coefs = list(); plots = list(); fits = list()
  # Remove each patient (LOO)
  for (pt in names(pg.samp)) {
    print(pt)
    tmp.patient.data = patient.data[subset(sum.patient.data, Patient != pt)$Patient]
    samples = as.vector(unlist(sapply(tmp.patient.data, function(df) df$info$Samplename )))

    train.rows = which(rownames(dysplasia.df) %in% samples)
    training = dysplasia.df[train.rows,]
    test = as.matrix(dysplasia.df[-train.rows,])
    if ( nrow(test) == ncol(dysplasia.df) ) test = t(test)

    # Predict function giving me difficulty when I have only a single sample, this ensures the dimensions are the same
    sparsed_test_data <- Matrix(data=0, nrow=ifelse(length(pg.samp[[pt]]$Samplename) > 1, nrow(test), 1),  ncol=ncol(training),
                                dimnames=list(pg.samp[[pt]]$Samplename,colnames(training)), sparse=T)
    for(i in colnames(dysplasia.df)) sparsed_test_data[,i] = test[,i]

@@

      pm = predict(fitLOO, newx=sparsed_test_data, s=cv$lambda.1se, type='response')
      sy = as.matrix(sqrt(binomial.deviance(pm, labels[pg.samp[[pt]]$Samplename])))

@@

    } else {
      warning(paste("Patient", pt, "did not have a 1se"))
    }
  }
  save(plots, performance.at.1se, coefs, fits, pg.samp, file=file)
}

@@

hgd = do.call(rbind, lapply(pg.samp[subset(sum.patient.data, Status == 'P')$Patient], function(pt) {
  hgd = subset(pt, Pathology %in% c('HGD', 'IMC'))
  cbind.data.frame('correctly.predicted'=sum(as.integer(hgd$Prediction > 0.5)), 'total.hgd.samples'=nrow(hgd))
}))
rownames(hgd) = subset(sum.patient.data, Status == 'P')$Patient
hgd = subset(hgd, total.hgd.samples > 0) # Some patients had no hgd samples

hgd$pred.ratio = hgd$correctly.predicted/hgd$total.hgd.samples

hgd$Patients = rownames(hgd)
hgd = transform(hgd, Patients=reorder(Patients, -pred.ratio) )

ggplot(hgd[,c('total.hgd.samples','pred.ratio', 'Patients')], aes(x=Patients, y=pred.ratio)) + 
  geom_bar(stat='identity', fill='darkblue', alpha=0.6) + geom_label( aes(label=total.hgd.samples) ) +
  coord_flip() + labs(title='Correctly predicted HGD samples', y='Correct prediction ratio')

lgd = do.call(rbind, lapply(pg.samp[subset(sum.patient.data, Status == 'P')$Patient], function(pt) {
  e = subset(pt, Pathology  == 'LGD')
  cbind.data.frame('predicted'=sum(as.integer(e$Prediction > 0.5)), 'total.lgd'=nrow(e))
}))
lgd = subset(lgd, total.lgd > 0)

lgd$pred.ratio = round(lgd$predicted/lgd$total.lgd, 2)

lgd$Patients = rownames(lgd)
lgd = transform(lgd, Patients=reorder(Patients, -pred.ratio) )

ggplot(lgd[,c('total.lgd','pred.ratio', 'Patients')], aes(x=Patients, y=pred.ratio)) + 
  geom_bar(stat='identity', fill='darkgreen', alpha=0.6) + geom_label( aes(label=total.lgd) ) +
  coord_flip() + labs(title='Correctly predicted LGD samples', y='Correct prediction ratio')

@@

all.predictions = do.call(rbind, lapply(all.predictions, function(x) {
  cbind.data.frame('correctly.predicted'=sum(x), 'total.samples'=length(x))
}))

all.predictions$Status = ifelse (rownames(all.predictions) %in% subset(sum.patient.data, Status == 'P')$Patient, 'P', 'NP')

#ggplot(melt(all.predictions, id.vars='Status'), aes(y=value, x=Status, fill=variable)) +
#  geom_bar(stat='identity', position='dodge') + labs(x='', y='Sample Counts', main='Total Sample Predictions')

adj.pred = as.data.frame(all.predictions %>% rowwise() %>% summarise( 
  'pred'=round(correctly.predicted/(sum(correctly.predicted, total.samples )),2),
  'total'=round(total.samples/(sum(correctly.predicted, total.samples )),2),
  'Status'=Status))
adj.pred$Patient = rownames(all.predictions)

adj.pred = transform(adj.pred, Patient=reorder(Patient, -total) )

m = melt(adj.pred, id.vars=c('Status', 'Patient'))
m = arrange(m, Status, variable, -value)
m$Status = as.factor(m$Status)

grid.arrange(
  ggplot(subset(m, Status == 'P') , aes(x=Patient, y=value, fill=variable)) + geom_bar(stat = 'identity') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + coord_flip() + labs(title='Progressors',x='', y='ratio'),
  ggplot(subset(m, Status == 'NP') , aes(x=Patient, y=value, fill=variable)) + geom_bar(stat = 'identity') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + coord_flip() + labs(title='Non-Progressors',x='', y='ratio')
)

pg = subset(adj.pred, Status == 'P')
np = subset(adj.pred, Status == 'NP')

@@

In progressors, `r round(sum(hgd$correctly.predicted)/sum(hgd$total.hgd.samples), 2)*100`% of the HGD samples were correctly predicted.

When predicting each sample individually `r round(nrow(subset(pg, pred == 0.5))/nrow(pg), 2)*100`% patients were correctly predicted in all samples. `r round(nrow(subset(pg, pred < 0.5 & pred > 0))/nrow(pg),2)*100`% of patients were correctly predicted in some samples

`r nrow(subset(pg, pred == 0))` patients had no correctly predicted samples, however, both had only two samples:
```{r, echo=F, warning=F, message=F}
x = do.call(rbind, lapply(pg.samp[subset(pg, pred == 0)$Patient], function(df) df[,c('Status','Patient','Path.ID','Endoscopy.Year', 'Pathology')] ))
row.names(x) = NULL
pander(x)
```

@@

In non-progressors `rnrow(subset(np, pred == 0.5))` patients were predicted to be non-progressors in _all_ of their samples, ranging from `r paste(range(all.predictions[subset(np, pred == 0.5)$Patient, 'total.samples']), collapse='-')`. `r nrow(subset(np, pred >= 0.3))` patients were predicted correctly in 70% or more of their samples. Only `r nrow(subset(np, pred < 0.3))` patients had fewer than 70% of their samples predicted correctly, and both of these have consistently been predicting _progression_ across multiple endoscopies.

`r pander(all.predictions[subset(np, pred < 0.3)$Patient,], caption='Non-progressors that have consistently predicted progression')`

```{r timepoints2, echo=T, message=F, warning=F}
# Only BE, Progressors
ndbe.only = do.call(rbind, lapply(pg.samp[subset(sum.patient.data, Status == 'P')$Patient], function(pt) {
  e = subset(pt, Pathology %in% c('BE', '', '?'))
  cbind.data.frame('predicted'=sum(as.integer(e$Prediction > 0.5)), 'total.be'=nrow(e))
}))
ndbe.only = subset(ndbe.only, total.be > 0)
ndbe.only$pred.ratio = round(ndbe.only$predicted/ndbe.only$total.be, 2)

ndbe.only$Patients = rownames(ndbe.only)
ndbe.only = transform(ndbe.only, Patients=reorder(Patients, -pred.ratio) )

ggplot(ndbe.only[,c('total.be','pred.ratio', 'Patients')], aes(x=Patients, y=pred.ratio)) + 
  geom_bar(stat='identity', fill='darkred', alpha=0.6) + geom_label( aes(label=total.be) ) +
  coord_flip() + labs(title='Correctly predicted NDBE samples', y='Correct prediction ratio')

total = do.call(rbind, lapply(pg.samp[subset(sum.patient.data, Status == 'P')$Patient], function(pt) {
    cbind.data.frame('predicted'=sum(as.integer(pt$Prediction > 0.5)), 'total'=nrow(pt))
}))

total$pred.ratio = round(total$predicted/total$total, 2)

total$Patients = rownames(total)
total = transform(total, Patients=reorder(Patients, -pred.ratio) ) 
ggplot(total[,c('total','pred.ratio', 'Patients')], aes(x=Patients, y=pred.ratio)) + 
  geom_bar(stat='identity', fill='purple', alpha=0.6) + geom_label( aes(label=total) ) +
  coord_flip() + labs(title='Correctly predicted Progressor samples', y='Correct prediction ratio')

# For all pre-HGD/IMC/LGD in progressors
pre.hgd = sapply(pg.samp[subset(sum.patient.data, Status == 'P')$Patient], function(pt) {
  e = subset(pt, Pathology %nin% c('HGD', 'IMC', 'LGD'))
  e$Prediction
  sum(as.integer(e$Prediction > 0.5))/nrow(e)
})

pred.by.endo<-function(pt, nyears) {
  e = subset(pt, PID %in% unique(pt$PID)[nyears])
  if (  length(which(e$Pathology %in% c('HGD','IMC'))) > 0 ) return(NA)
  p = as.vector(as.integer(e$Prediction > 0.5))
  names(p) = e$Endoscopy.Year
  return(p)
}

@@

cols = c('%.Predicted','n.Endoscopies','n.Patients', 'FNR','FPR', 'median.pred.samp','median.total.samp', 'max.samples', 'sd.samples')
p.predictions = as.data.frame(matrix(data=0,nrow=6, ncol=length(cols), dimnames=list(c(), cols)))
for (i in 1:nrow(p.predictions)) {
  p = sapply(pg.samp[subset(sum.patient.data, Status == 'P')$Patient], pred.by.endo, i)
  pp = sapply(p, predicted.progression)
  pp = pp[!is.na(pp)]

  fpr = sapply(pg.samp[subset(sum.patient.data, Status == 'NP')$Patient], pred.by.endo, i)
  fpr = sapply(fpr, predicted.progression)
  fpr = fpr[!is.na(fpr)]

  p.predictions[i,] =  rbind(round(sum(pp[!is.na(pp)]/length(pp[!is.na(pp)])), 2), i, length(pp), 
                             round(length(which(pp == 0))/length(pp), 2),
                             round(sum(fpr)/length(fpr), 2), 
                             median(sapply(p[names(which(pp > 0) )], function(x) length(which(x == 1)) )),
                             median(sapply(p[names(which(pp > 0))], length)), 
                             max(sapply(p[names(which(pp > 0))], length)),
                             sd(sapply(p[names(which(pp > 0))], length))
                             )
}

ggplot(p.predictions, aes(y=`%.Predicted`,x=factor(n.Endoscopies))) + 
  geom_bar(stat="identity", fill='dodgerblue3') + 
  geom_text(aes(label=paste(`%.Predicted`*100, '%', sep='')), nudge_y=-0.02, size=8) + 
  geom_text(aes(label=paste(n.Patients, 'patients')), nudge_y=0.01, size=8) + 
  geom_bar(stat='identity', fill='darkred', alpha=0.3, aes(y=FPR)) +
  geom_text(aes(label=FPR, y=FPR)) +
  labs(title="Progression predicted per endoscopy", x='Number of Endoscopies', y='% Patients') +
  theme(text = element_text(size = 14))

@@

## What clinicicians really want

Backwards from the final sample.

```{r, warning=F, message=F, echo=F, fig.width=16, fig.height=8}

back = lapply(pg.samp[subset(sum.patient.data, Status == 'P')$Patient], function(pt) {
  pt = ddply(pt, .(PID, Endoscopy.Year), summarise, 'pred'=sum(as.integer(Prediction > 0.5)), 'samples'=length(PID), 
        'final.endo'= length(which(grepl('HGD|IMC', Pathology))) > 0 )
  arrange(pt, -Endoscopy.Year, PID)
})

endos = do.call(rbind, lapply(1:max(sapply(back,nrow)), function(i) {
  df = do.call(rbind, lapply(back, function(x) {
    if (i > nrow(x)) return(NA)
    cbind(x[i, c('Endoscopy.Year','pred','samples','final.endo')], i)
  }))
  df$Patient = rownames(df)
  return(df)
}))

endos = endos[complete.cases(endos),]
endos$i = endos$i-1

endos$pred.ratio = round(endos$pred/endos$samples, 2)
endos$i = as.factor(endos$i)
